Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-11-18 _tex/index.tex
Built site for gh-pages

However, Case V (and its BTL model counterpart) was originally developed
to provide a ``rather coarse scaling'' of traits
\citep[pp.~269]{Thurstone_1927b}, prioritizing statistical simplicity
over precision in trait measurement \citep[pp.~677]{Kelly_et_al_2022}.
As a result, its assumptions may not suit applications beyond the
psycho-physical contexts for which it was created. Thurstone himself
cautioned that its use ``should not be made without (an) experimental
test'' \citep[pp.~270]{Thurstone_1927b}, acknowledging that some
assumptions could prove problematic with more complex, less homogeneous
stimuli, such as handwriting or English compositions
\citep[pp.~374]{Thurstone_1927a}. Consequently, given that current CJ
applications often deal with complex, less homogeneous stimuli, two key
assumptions of Case V may not consistently hold in theory or practice:
the zero correlation and equal dispersion between stimuli.

The assumption of \emph{zero correlation between stimuli} is best
illustrated with an example. For instance, when evaluating text quality,
this assumption suggests that a judge's perception of quality in one
text does not influence their perception of the same trait in the
comparison text. Thurstone attributes this independence to the
cancellation of potential judges' biases, driven by opposing and equally
weighted factors; commonly known as `mood' and `simultaneous contrast'
effects, that operate during pairwise comparisons
\citep[pp.~268]{Thurstone_1927b}. This cancellation of bias has been
mathematically demonstrated by \citet{Andrich_1978}, assuming additive
biases and a logit scale derived from the BTL model.

However, given that evidence has shown judges' quality assessments are
influenced by multiple, complex aspects of the texts
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}, it is
plausible that the accuracy of these judgments also varies based on
factors such as age, culture, education, expertise
\citep[pp.~683]{Kelly_et_al_2022}, and ultimately, based on the
individual judge themselves
\citep{Gill_et_al_2013, vanDaal_et_al_2017, vanDaal_2020}.

{This means that the cancellation of judges' biases may not hold in
theory or practice, and that judges' biases may depend on the stimuli,
by means of an interaction (but present) or because the judges works as
a confounder}

>>>>

However, Case V was originally developed to provide a ``rather coarse
scaling'' of traits \citep[pp.~269]{Thurstone_1927b}, prioritizing
statistical simplicity over precision in trait measurement
\citep[pp.~677]{Kelly_et_al_2022}. As a result, its assumptions may not
suit applications beyond the psycho-physical contexts for which it was
created. Thurstone himself cautioned that its use ``should not be made
without experimental test'' \citep[pp.~270]{Thurstone_1927b},
acknowledging that some assumptions could prove problematic with complex
traits or less homogeneous stimuli \citep[pp.~374]{Thurstone_1927a}.
Consequently, given that current CJ applications often deal these types
of traits and stimuli, two key assumptions of Case V may not
consistently hold in theory or practice: the zero correlation and equal
dispersion between stimuli.

The assumption of \emph{zero correlation between stimuli} is best
illustrated with an example. For instance, when evaluating text quality,
the assumption suggests that a judge's perception of quality in one text
does not influence the perception of the same trait in the comparison
text. Thurstone attributes this independence to the cancellation of
potential judges' biases, driven by opposing and equally weighted
factors that operate during pairwise comparisons, called `mood' and
`simultaneous contrast' effects \citep[pp.~268]{Thurstone_1927b}. This
cancellation of bias has been mathematically demonstrated by
\citet{Andrich_1978}, assuming discriminal processes with additive
biases and a logit scale derived from the BTL model.

However, two types of scenarios make it plausible that potential judges'
biases may not cancel: those involving complex traits and less
homogeneous stimuli, and those where judges differ in what they value in
their assessments. In the first scenario, Thurstone noted that for
complex traits and non-homogeneous stimuli, such as handwriting or
English compositions, CJ data might not align with the assumptions of
Case V \citep[pp.~374]{Thurstone_1927a}. This insight likely extends to
other similarly complex traits and stimuli. In the second scenario,
evidence indicates that judges' assessments are influenced by multiple,
intricate aspects of the stimuli
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
Moreover, factors like age, culture, education, expertise
\citep[pp.~683]{Kelly_et_al_2022}, and even individual differences among
judges \citep{Gill_et_al_2013, vanDaal_et_al_2017, vanDaal_2020} can
influence judgment accuracy. These scenarios can ultimately result in
non-additive biases that resist cancellation, driven by the
characteristics or location of stimuli within the trait continuum.

{This also translates into the idea that stimuli are the main focus of
estimation and analysis, but what happens when the focus of analysis is
the individuals that generated those stimuli. Meaning there is an amount
of correlation that it is not accounted for. Use example of Boonen!!}
